On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-06 Manuscript/manuscript.tex
Worked on table 3

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the sun goes off.  Thus, we can try
to reduce this variability by accounting for this in the model through the batch
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
\eds{I was talking to my husband about this. Could the heat effect be justified 
to capture the variability in the amount of time between the ``set'' and the 
gun going off (assuming this is different from heat to heat)? E.g., quicker 
gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
but this effect would be shared across all athletes in the same heat.}
\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

\begin{table}
  \centering
  \caption{Summary statistics for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Excluding 2022 & $-1912.9$ &  0.058 & 0.033 & $-1.868$ & 0.136 \\
      All Data & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

>>>>

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for this in the model through the batch
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
\eds{I was talking to my husband about this. Could the heat effect be justified 
to capture the variability in the amount of time between the ``set'' and the 
gun going off (assuming this is different from heat to heat)? E.g., quicker 
gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
but this effect would be shared across all athletes in the same heat.}
\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Year Effect Only & $-2023.5$ & N/A & 0.048 & $-1.857$ & 0.152
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135
      Full model & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}
